---
title: "Detecting human-mouse multiplets in PDX single-cell data with multipletR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting human-mouse multiplets in PDX single-cell data with multipletR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 4
)
```

## Overview

In a patient-derived xenograft (PDX) experiment a human tumor grows inside a
mouse host, so a dissociated sample contains both human tumor cells and mouse
stromal and immune cells. When such a sample is run through droplet-based
single-cell RNA-sequencing, some droplets capture a human cell and a mouse cell
together. These cross-species **multiplets** must be found and removed before
downstream analysis.

`multipletR` detects these multiplets from the file 10x Cell Ranger already
produces, using an adaptive threshold method that, unlike a fixed cutoff, does
not assume a fixed human/mouse proportion. This vignette walks through the input
it needs, the main detection function, the diagnostic plots, and how to use the
results with a Seurat object.

```{r load}
library(multipletR)
```

## The input: Cell Ranger's GEM classification file

The package takes a single file as its primary input: the GEM classification CSV
produced by Cell Ranger when reads are aligned to a combined two-species
reference. This section explains where that file comes from, where to find it,
and what it contains.

### How the data is generated

In droplet-based single-cell RNA-sequencing (10x Genomics Chromium), each
droplet (a GEM, Gel Bead-in-EMulsion) captures one or more cells together with a
barcoded gel bead, so every cell's transcripts inherit a shared cell barcode and
each molecule gets a unique molecular identifier (UMI). Most droplets contain a
single cell (a singlet), but some capture two or more; when those cells come from
different species, the droplet is a human-mouse multiplet.

To tell the two species apart, reads are aligned with `cellranger count`
against a combined ("barnyard") reference that contains both genomes, for example
GRCh38 (human) and GRCm39 (mouse). In this reference every gene is prefixed by
its genome of origin (e.g. `GRCh38-EPCAM`, `GRCm39-Col1a1`), so each read can be
attributed to human or mouse. For every cell barcode, Cell Ranger counts how
many reads map to each genome and, using fixed count thresholds, labels the
barcode as human, mouse, or a multiplet. This per-barcode, two-genome summary is
written to the GEM classification file.

### Where to find it

The file is written only when the reference is a multi-genome (barnyard)
reference. For a run with `--id=<SAMPLE>`, it is located at:

```
<SAMPLE>/outs/analysis/gem_classification.csv
```

The full expression matrix lives separately at
`<SAMPLE>/outs/filtered_feature_bc_matrix/`. `multipletR` does not need the
matrix, only `gem_classification.csv`, although the Seurat helper later joins the
calls back onto a matrix-derived object by barcode.

### What it contains

One row per cell barcode, with these columns:

| Column | Meaning |
|---|---|
| `barcode` | The 10x cell barcode (often carries a `-1` suffix, e.g. `AAACCTGAG...-1`). |
| *human genome* | Per-barcode count of reads assigned to the human genome, named after the reference build, e.g. `GRCh38`. |
| *mouse genome* | Per-barcode count of reads assigned to the mouse genome, named after the build: `GRCm39` (2024-A references), or `mm10` / `mm39` for older builds. |
| `call` | Cell Ranger's own classification: the human genome name (e.g. `GRCh38`), the mouse genome name (e.g. `GRCm39`), or `Multiplet`. |

The package ships a real PDX example (sample 12G). Here are the first rows:

```{r peek-input}
gem_file <- system.file("extdata", "12G_gem_classification.csv",
                        package = "multipletR")
head(read.csv(gem_file))
```

`multipletR` derives, for each barcode, the total reads (human + mouse) and the
percent mouse (mouse / total x 100), and applies its adaptive thresholds in that
total-reads by percent-mouse space rather than trusting Cell Ranger's fixed
`call`. The `call` column is kept only for the comparison plots.

### Column-name flexibility

Because the genome columns are named after whichever reference build was used,
`multipletR` recognizes several common names automatically, so you do not have to
rename anything:

- **Human:** `GRCh38`, `hg38`, `human_reads`, `human`
- **Mouse:** `GRCm39`, `mm10`, `mm39`, `mouse_reads`, `mouse`
- **Barcode:** `barcode`, `Barcode`, `barcodes` (if none is present, the row index is used as a fallback)
- **Call:** `call`, `Call`, `classification`

As long as the file has a human read-count column and a mouse read-count column,
`detect_multiplets()` will run; the barcode and call columns are optional but
recommended (the barcode is what lets you match results back to a Seurat object).

**Tip:** barcode suffixes must match between the GEM file and any Seurat object
you later filter. Cell Ranger writes barcodes with a trailing `-1`; if your
Seurat object's cell names lack it (or vice versa), `remove_multiplets_seurat()`
will match nothing. Make the two consistent before joining.

## Detecting multiplets

`detect_multiplets()` reads the GEM classification file, runs the adaptive
threshold engine, writes an annotated CSV, and (by default) draws diagnostic
plots. It returns the input data with three added columns: `our_classification`
(`Multiplet` / `Singlet`), `pct_human`, and `pct_mouse`.

```{r detect, fig.alt = "Diagnostic percent plots colored by 10X and by our classification"}
res <- detect_multiplets(
  fileIn  = gem_file,
  fileOut = tempfile(fileext = ".csv")
)
```

The one-line summary reports the final thresholds and the number of multiplets
found. The returned data frame carries the per-cell result:

```{r result}
head(res)

table(res$our_classification)
```

### Reading the diagnostic plot

The plot drawn above shows total reads (x) against percent mouse (y), with two
panels: the left colored by Cell Ranger's `call`, the right by our
classification. In the right panel our multiplets sit in the central
percent-mouse band, where a droplet carries a real mix of human and mouse reads.
Pure human cells fall near 0% mouse and pure mouse cells near 100%, so the
multiplets we keep are the balanced ones in between.

We can confirm the multiplets are genuinely balanced by looking at their
composition, which centers near a 50/50 human/mouse mix:

```{r composition}
mult <- res[res$our_classification == "Multiplet", ]
summary(mult$pct_human)
summary(mult$pct_mouse)
```

### Adjusting the thresholds

The defaults are the recommended values and rarely need changing, but every
parameter is adjustable. The three starting thresholds are `T1` (upper percent
mouse), `T2` (lower percent mouse), and `T3` (lower total-reads percentile); the
two stopping sensitivities are `overlapDrop` and `modeDiff`. For example, a more
conservative run:

```{r adjust, eval = FALSE}
res_strict <- detect_multiplets(
  fileIn      = gem_file,
  fileOut     = tempfile(fileext = ".csv"),
  overlapDrop = 5,      # stop expanding sooner
  modeDiff    = 0.5     # stricter on distribution similarity
)
```

See `?detect_multiplets` for the full list of arguments.

## Using the results with Seurat

Once multiplets are detected, `remove_multiplets_seurat()` works directly on a
Seurat object. It adds the per-cell classification to the object's metadata
(`multipletR_class`, `multipletR_pct_human`, `multipletR_pct_mouse`) and, by
default, removes the detected multiplets so the object is ready for downstream
analysis. Setting `remove = FALSE` keeps all cells and only annotates them, which
is useful for visualizing where the multiplets fall (for example on a UMAP
colored by `multipletR_class`) before deciding whether to filter.

```{r seurat, eval = FALSE}
library(Seurat)

# Build a Seurat object from the same sample's count matrix, then:

# Annotate only (keep all cells) to inspect the multiplets first
seu <- remove_multiplets_seurat(seu, res, remove = FALSE)
table(seu$multipletR_class)
DimPlot(seu, group.by = "multipletR_class")

# Or annotate and remove in one step, ready for downstream analysis
seu_clean <- remove_multiplets_seurat(seu, res)
```

The Seurat steps are shown but not run here, since they need the full count
matrix rather than just the GEM classification file.

## Summary

`multipletR` centers on two functions. `detect_multiplets()` takes a Cell Ranger
GEM classification file and adds an adaptive multiplet classification plus the
diagnostic plots; `remove_multiplets_seurat()` carries that result onto a Seurat
object, annotating and optionally removing the multiplets. Because the method
adapts its thresholds to each sample rather than assuming a fixed species
proportion, it handles the imbalanced human/mouse mixtures typical of real PDX
data.
