---
title: "Detecting human-mouse multiplets in PDX single-cell data with multipletR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting human-mouse multiplets in PDX single-cell data with multipletR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4
)
```

## Overview

In a patient-derived xenograft (PDX) experiment a human tumor grows inside a
mouse host, so a dissociated sample contains both human tumor cells and mouse
stromal and immune cells. When such a sample is run through droplet-based
single-cell RNA-sequencing, some droplets capture a human cell and a mouse cell
together. These cross-species **multiplets** must be found and removed before
downstream analysis.

`multipletR` detects these multiplets from the file 10x Cell Ranger already
produces, using an adaptive threshold method that, unlike a fixed cutoff, does
not assume a fixed human/mouse proportion. This matters in real PDX samples,
where the human/mouse balance shifts from sample to sample and a fixed cutoff
tends to over-call multiplets. This vignette walks through the input it needs,
the main detection function, the diagnostic plots, and how to use the results
with a Seurat or SingleCellExperiment object.

```{r load}
library(multipletR)
```

## The input: Cell Ranger's GEM classification file

The package takes a single file as its primary input: the GEM classification CSV
produced by Cell Ranger when reads are aligned to a combined two-species
reference. This section explains where that file comes from, where to find it,
and what it contains.

### How the data is generated

In droplet-based single-cell RNA-sequencing (10x Genomics Chromium), each
droplet (a GEM, Gel Bead-in-EMulsion) captures one or more cells together with a
barcoded gel bead, so every cell's transcripts inherit a shared cell barcode and
each molecule gets a unique molecular identifier (UMI). Most droplets contain a
single cell (a singlet), but some capture two or more; when those cells come from
different species, the droplet is a human-mouse multiplet.

To tell the two species apart, reads are aligned with `cellranger count`
against a combined ("barnyard") reference that contains both genomes, for example
GRCh38 (human) and GRCm39 (mouse). In this reference every gene is prefixed by
its genome of origin (e.g. `GRCh38-EPCAM`, `GRCm39-Col1a1`), so each read can be
attributed to human or mouse. For every cell barcode, Cell Ranger counts how
many reads map to each genome and, using fixed count thresholds, labels the
barcode as human, mouse, or a multiplet. This per-barcode, two-genome summary is
written to the GEM classification file.

### Where to find it

The file is written only when the reference is a multi-genome (barnyard)
reference. For a run with `--id=<SAMPLE>`, it is located at:

```
<SAMPLE>/outs/analysis/gem_classification.csv
```

The full expression matrix lives separately at
`<SAMPLE>/outs/filtered_feature_bc_matrix/`. `multipletR` does not need the
matrix, only `gem_classification.csv`, although the Seurat/SingleCellExperiment
helper later joins the calls back onto a matrix-derived object by barcode.

### What it contains

One row per cell barcode, with these columns:

| Column | Meaning |
|---|---|
| `barcode` | The 10x cell barcode (often carries a `-1` suffix, e.g. `AAACCTGAG...-1`). |
| *human genome* | Per-barcode count of reads assigned to the human genome, named after the reference build, e.g. `GRCh38`. |
| *mouse genome* | Per-barcode count of reads assigned to the mouse genome, named after the build: `GRCm39` (2024-A references), or `mm10` / `mm39` for older builds. |
| `call` | Cell Ranger's own classification: the human genome name (e.g. `GRCh38`), the mouse genome name (e.g. `GRCm39`), or `Multiplet`. |

A typical file looks like:

```
barcode,GRCh38,GRCm39,call
AAACCTGAGAAACCAT-1,10432,58,GRCh38
AAACCTGAGATCCGAG-1,71,9903,GRCm39
AAACCTGCACGTGTGA-1,4821,5210,Multiplet
```

The first row is dominated by human reads (a human singlet), the second by mouse
reads (a mouse singlet), and the third has substantial reads from both genomes (a
multiplet).

The package ships a real PDX example (sample PC65). Here are the first rows:

```{r peek-input}
gem_file <- system.file("extdata", "PC65_gem_classification.csv",
  package = "multipletR"
)
head(read.csv(gem_file))
```

### Why this is the input multipletR needs

The package's whole premise is that a true human-mouse multiplet has substantial
reads from both genomes, whereas a singlet is dominated by one. The GEM
classification file provides exactly the two numbers needed to test that: the
human and mouse read counts per barcode. From them, `multipletR` derives, for
each barcode, the total reads (human + mouse) and the percent mouse
(mouse / total x 100), and applies its adaptive thresholds in that
total-reads by percent-mouse space rather than trusting Cell Ranger's fixed
`call`. The `call` column is kept only for the comparison plots.

### Column-name flexibility

Because the genome columns are named after whichever reference build was used,
`multipletR` recognizes several common names automatically, so you do not have to
rename anything:

- **Human:** `GRCh38`, `hg38`, `human_reads`, `human`
- **Mouse:** `GRCm39`, `mm10`, `mm39`, `mouse_reads`, `mouse`
- **Barcode:** `barcode`, `Barcode`, `barcodes` (if none is present, the row index is used as a fallback)
- **Call:** `call`, `Call`, `classification`

As long as the file has a human read-count column and a mouse read-count column,
`detect_multiplets()` will run; the barcode and call columns are optional but
recommended (the barcode is what lets you match results back to a Seurat or
SingleCellExperiment object).

**Tip:** barcode suffixes must match between the GEM file and any object you
later filter. Cell Ranger writes barcodes with a trailing `-1`; if your object's
cell names lack it (or vice versa), `remove_multiplets()` will match nothing.
Make the two consistent before joining.

## Detecting multiplets

`detect_multiplets()` reads the GEM classification file, runs the adaptive
threshold engine, writes an annotated CSV, and (by default) draws diagnostic
plots. It returns the input data with three added columns: `our_classification`
(`Multiplet` / `Singlet`), `pct_human`, and `pct_mouse`.

Here we draw both diagnostic views: the percent plot (total reads vs percent
mouse) and the total-reads plot (mouse reads vs human reads).

```{r detect, fig.alt = "Diagnostic plots colored by 10X and by multipletR classification"}
res <- detect_multiplets(
  fileIn  = gem_file,
  fileOut = tempfile(fileext = ".csv"),
  plotPercent    = TRUE,
  plotTotalReads = TRUE
)
```

The one-line summary reports the final thresholds and the number of multiplets
found. The returned data frame carries the per-cell result:

```{r result}
head(res)
table(res$our_classification)
```

### Reading the diagnostic plots

Each plot has two panels: the left colored by Cell Ranger's `call`, the right by
the multipletR classification. Across both panels human singlets are blue and
mouse singlets are green; Cell Ranger's multiplets are orange (left) and
multipletR's are red (right), so it is easy to see which cells each method flags.

In the percent plot, multiplets sit in the central percent-mouse band, where a
droplet carries a real mix of human and mouse reads; pure human cells fall near
0% mouse and pure mouse cells near 100%. In the total-reads plot, true multiplets
have substantial reads from *both* genomes, so they sit away from the two axes.

The contrast with Cell Ranger is the key point. In this PDX sample Cell Ranger's
fixed thresholds label many predominantly single-species barcodes as multiplets,
whereas the adaptive method keeps only the balanced droplets. You can see the
size of that gap directly:

```{r compare-counts}
# Cell Ranger's multiplet calls (from the GEM file)
sum(read.csv(gem_file)$call == "Multiplet")
# multipletR's multiplet calls
sum(res$our_classification == "Multiplet")
```

We can confirm the multiplets multipletR keeps are genuinely balanced by looking
at their composition, which centers near a mixed human/mouse split rather than
sitting at one extreme:

```{r composition}
mult <- res[res$our_classification == "Multiplet", ]
summary(mult$pct_human)
summary(mult$pct_mouse)
```

### Adjusting the thresholds

The defaults are the recommended values and rarely need changing, but every
parameter is adjustable. The three starting thresholds are `T1` (upper percent
mouse), `T2` (lower percent mouse), and `T3` (lower total-reads percentile); the
two stopping sensitivities are `overlapDrop` and `modeDiff`. For example, a more
conservative run:

```{r adjust, eval = FALSE}
res_strict <- detect_multiplets(
  fileIn      = gem_file,
  fileOut     = tempfile(fileext = ".csv"),
  overlapDrop = 5, # stop expanding sooner
  modeDiff    = 0.5 # stricter on distribution similarity
)
```

See `?detect_multiplets` for the full list of arguments.

## Using the results with Seurat or SingleCellExperiment

Once multiplets are detected, `remove_multiplets()` carries the result onto a
single-cell object. It adds the per-cell classification to the object's cell
metadata (`multipletR_class`, `multipletR_pct_human`, `multipletR_pct_mouse`)
and, by default, removes the detected multiplets so the object is ready for
downstream analysis. Setting `remove = FALSE` keeps all cells and only annotates
them, which is useful for visualizing where the multiplets fall (for example on a
UMAP colored by `multipletR_class`) before deciding whether to filter.

Use `object = "seurat"` for a Seurat object or `object = "sce"` for a
`SingleCellExperiment`; the classification logic is identical, only the metadata
is written to the appropriate slot.

```{r seurat, eval = FALSE}
library(Seurat)
# Build a Seurat object from the same sample's count matrix, then:
# Annotate only (keep all cells) to inspect the multiplets first
seu <- remove_multiplets(seu, res, object = "seurat", remove = FALSE)
table(seu$multipletR_class)
DimPlot(seu, group.by = "multipletR_class")
# Or annotate and remove in one step, ready for downstream analysis
seu_clean <- remove_multiplets(seu, res, object = "seurat")
```

The same call works for a `SingleCellExperiment`:

```{r sce, eval = FALSE}
library(SingleCellExperiment)
# With an existing SingleCellExperiment 'sce' whose colnames are barcodes:
sce <- remove_multiplets(sce, res, object = "sce", remove = FALSE)
table(sce$multipletR_class)
# Or annotate and remove in one step
sce_clean <- remove_multiplets(sce, res, object = "sce")
```

The object steps are shown but not run here, since they need the full count
matrix rather than just the GEM classification file.

## Summary

`multipletR` centers on two functions. `detect_multiplets()` takes a Cell Ranger
GEM classification file and adds an adaptive multiplet classification plus the
diagnostic plots; `remove_multiplets()` carries that result onto a Seurat or
SingleCellExperiment object, annotating and optionally removing the multiplets.
Because the method adapts its thresholds to each sample rather than assuming a
fixed species proportion, it handles the imbalanced human/mouse mixtures typical
of real PDX data, where a fixed cutoff over-calls multiplets.

## Session info

```{r session-info}
sessionInfo()
```
